---
title: Individual Claim Reserving
engine: knitr
# Only utilities.js and the esbuild-compiled components ship to docs/; the
# .jsx sources stay in the repo as the editable source of truth (compiled by
# scripts/compile_interactive_splits.py, invoked from this file's python
# setup chunk so it runs only when this lecture renders).
resources:
  - interactive-splits/utilities.js
  - interactive-splits/compiled/*.js
# The project-wide `cache: true` must be off here: knitr's chunk cache does
# not replay python side effects, so later chunks see an empty session.
# Cheap anyway: SPLICE generation is only ~2.5s.
execute:
  cache: false
bibliography: ../library.bib
format-links:
  - html
  - format: revealjs
    text: Slides (html)
  - text: Slides (pdf)
    href: individual-claim-reserving.pdf
    icon: file-pdf
---

<!--
engine: knitr (not the course-wide jupyter) so R and Python share one file:
R chunks run natively (SPLICE, ChainLadder); Python runs via reticulate ->
../.venv. knitr gotchas that cause real breakage:
- cache: false in the front matter — knitr's cache doesn't replay python
  state, so a cached edited chunk runs in an empty session (NameError).
- Data generation is hoisted to the setup chunk at the top: it must run before
  any python chunk reads data/splice/*.parquet, and rewriting an already-read
  parquet mid-render corrupts later reads (pyarrow mmap). Don't move it down or
  merge it into the case-study slide (which shows the same code, eval: false).
  Reproducible from random_seed, ~2.5s, so the parquet is a build artifact.
- jupyter_compat = TRUE (first r chunk): only the last expression auto-prints.
  Never write `_ = ...` in a python chunk — it shadows reticulate's repr hook.
- #| fig-dpi: 350 on every python plotting chunk (knitr ignores rcParams dpi).
- A bare dataframe renders as plain text; use show_df() for HTML tables.
- Bind/chain reticulate results (glm = make_glm().fit()) so knitr doesn't emit
  sklearn's HTML estimator diagram and swallow the chunk output.
-->

```{r}
#| echo: false
#| warning: false
# Bind reticulate to the project venv BEFORE any python chunk runs.
# (Package loading happens on the case study's imports slide.)
reticulate::use_virtualenv("../.venv", required = TRUE)
# Display python values like the jupyter engine: only the final expression
# in a chunk is auto-printed (a trailing `;` still suppresses it), instead
# of reticulate's default REPL-style echo of every top-level expression.
# And centre figures horizontally (knitr's default is left-aligned,
# unlike the jupyter engine).
knitr::opts_chunk$set(jupyter_compat = TRUE, fig.align = "center")
```

```{r}
#| echo: false
#| output: false
# Data generation for the whole lecture, hoisted here so it runs before any
# python chunk reads data/splice/*.parquet. Don't move it down or rewrite the
# parquet later in the render: once pyarrow has mmapped a file, an in-place
# rewrite makes later reads hand back garbage. Shown again (eval: false) at
# the case study slide.
suppressMessages(library(SPLICE))
suppressMessages(library(arrow))
dir.create("data/splice", recursive = TRUE, showWarnings = FALSE)
# 100 quarters of history so the book reaches a steady state; we later keep
# only the claims finalising in the last ~40 quarters (see T_START below),
# mimicking a portfolio inherited mid-stream rather than a brand-new insurer.
sim <- generate_data(
  n_claims_per_period = 100, n_periods = 100, complexity = 5,
  random_seed = 20260712,
  covariates_obj = SynthETIC::test_covariates_obj
)
write_parquet(cbind(sim$claim_dataset, sim$covariates_data$data),
              "data/splice/claims.parquet")
write_parquet(sim$payment_dataset, "data/splice/payments.parquet")
write_parquet(sim$incurred_dataset, "data/splice/estimates.parquet")
```

```{python}
#| echo: false
#| warning: false
from setup import *

# Compile interactive-splits/*.jsx to compiled/*.js before quarto copies the
# `resources:`. No-op when already up to date.
import subprocess
import sys

subprocess.run(
    [sys.executable, "../scripts/compile_interactive_splits.py"], check=True
);


def show_df(df):
    """Emit a dataframe's HTML in an `output: asis` chunk, wrapped in the
    .cell-output div the jupyter engine uses (gives wide tables a scrollbar)."""
    print('<div class="cell-output cell-output-display">'
          + df._repr_html_() + "</div>")
```

---

::: {.callout-note}
These slides are currently a work in progress.
The full/final version will be completed soon.
:::

# Introduction

## Individual claim reserving

An accident has occurred, the insurer has been notified, and maybe parts of the claim have already been paid, now we need to guess how much remains in this realised claim.

```{=html}
<!-- The live version of the timeline that images/individual-claim.png was a
screenshot of. Mounted by lecture-demos.js (id demo-claim-intro), which loads
on the "One claim's full history" slide below; type="module" scripts run after
the whole document is parsed, so this earlier div is found and mounted fine.
The .interactive-demo CSS in that same later block applies globally. -->
<div id="demo-claim-intro" class="interactive-demo"></div>
```

This is potentially a promising application for neural networks.

## Ultimate claim size

For claim $k$, we want to know the  _ultimate claim size_
$$
  U_k := \sum_{t} P_{k,t},
$$

This is the sum of all _incremental payments_ $P_{k,t}$ (also called _partial payments_) over the complete lifetime of the claim.

::: {.callout-note}
## Assumption: payments are positive

We assume $P_{k,t} \ge 0$ throughout. In reality some payments are _negative_ --- called _recoveries_ (e.g. money coming back from a third party or a salvaged vehicle). They are typically so rare and small relative to normal payments that ignoring them is reasonable (though this depends on the dataset).
:::

## Outstanding claim amount

Say the valuation date is $v$, then we've seen the history of the claim $\mathcal{H}_{k,v}$ to that time, including past payments.

So the _outstanding_ (future unpaid amount) for claim $k$ is
$$
  O_k(v) := \sum_{t>v} P_{k,t}.
$$

Writing $P_k(v) := \sum_{t \le v} P_{k,t}$ for the _cumulative paid to date_, if the claim is still ongoing then
$$
  U_k = P_k(v) + O_k(v) \,.
$$

In words:
$$
\text{Ultimate} = \text{Cumulative payments before } v + \text{Outstanding amount from } v \text{ onwards}
$$

## Total outstanding

For a portfolio, we'd like to know the _total outstanding_

$$
  T(v) := \sum_{k\in\mathcal{O}(v)} O_k(v) \,,
$$
where $\mathcal{O}(v)$ is the set of all open claims at valuation time $v$.

Note: this doesn't include _incurred but not reported (IBNR)_ claims.

## The total outstanding, visually

```{python}
#| echo: false
#| fig-dpi: 350
# Hand-crafted schematic (not SPLICE data): circle = notification, x = payment,
# big X = finalisation. Open claims' post-v segments are highlighted (their
# future payments are the O_k(v)'s summing to T(v)); an IBNR claim at the bottom.
_v = 10
_diagram_claims = [  # (notification, [payment times], finalisation)
    (1.0, [2.5, 4.0], 5.5),
    (2.0, [3.0, 6.0, 8.5], 9.4),
    (3.5, [5.0, 8.0, 11.0, 13.0], 14.0),
    (5.0, [6.5, 9.0], 9.8),
    (6.0, [7.5, 12.0, 15.5], 17.0),
    (8.0, [9.5, 11.5, 14.0], 15.0),
]
_GREY, _TEAL, _PINK = "#B8B8B8", "#3F9999", "#FF8FA9"

plt.figure(figsize=(5, 2.8))
for _row, (_n, _pays, _f) in enumerate(_diagram_claims):
    _y = -_row
    if _n <= _v < _f:  # open at the valuation date
        plt.plot([_n, _v], [_y, _y], color=_TEAL, linewidth=1.5)
        plt.plot([_v, _f], [_y, _y], color=_PINK, linewidth=2.5)
        for _p in _pays:
            _c = _TEAL if _p <= _v else _PINK
            plt.plot(_p, _y, "x", color=_c, markersize=5)
        plt.plot(_n, _y, "o", color=_TEAL, markersize=4)
        plt.plot(_f, _y, "X", color=_PINK, markersize=6)
    else:  # already closed: greyed out
        plt.plot([_n, _f], [_y, _y], color=_GREY, linewidth=1.5)
        for _p in _pays:
            plt.plot(_p, _y, "x", color=_GREY, markersize=5)
        plt.plot(_n, _y, "o", color=_GREY, markersize=4)
        plt.plot(_f, _y, "X", color=_GREY, markersize=6)

# The IBNR claim: occurred before v, notified after v — invisible at v.
_y_ibnr = -len(_diagram_claims) - 0.6
plt.plot(8.5, _y_ibnr, "^", color="#666666", markersize=5)
plt.plot([8.5, 12.0], [_y_ibnr, _y_ibnr], color=_GREY,
         linewidth=1.2, linestyle=":")
plt.plot([12.0, 16.0], [_y_ibnr, _y_ibnr], color=_GREY, linewidth=1.5)
plt.plot(12.0, _y_ibnr, "o", color=_GREY, markersize=4)
for _p in [13.0, 15.0]:
    plt.plot(_p, _y_ibnr, "x", color=_GREY, markersize=5)
plt.plot(16.0, _y_ibnr, "X", color=_GREY, markersize=6)
plt.annotate("IBNR: accident before $v$, reported after",
             (8.2, _y_ibnr), ha="right", va="center", fontsize=7,
             color="#666666")

# Stop the dashed line below the label so the text doesn't sit on it.
plt.plot([_v, _v], [_y_ibnr - 0.9, 0.55], color="black",
         linestyle="--", linewidth=1)
plt.annotate("valuation time $v$", (_v, 0.9), ha="center",
             va="bottom", fontsize=8)
plt.annotate("$O_k(v)$", (12.0, -1.6), ha="center", va="bottom",
             fontsize=8, color=_PINK)
plt.annotate("$T(v)$ = sum of the\nhighlighted future payments",
             (17.2, -3.0), ha="right", va="center", fontsize=7)

plt.ylim(_y_ibnr - 0.9, 1.6)
plt.yticks([])
plt.gca().spines["left"].set_visible(False)
plt.xlabel("Time (quarters)");
```

Each open claim's future payments (highlighted) get summed to give $T(v)$.

The bottom claim has _occurred_ before $v$ but hasn't been _notified_ yet.
No individual model can reserve for a claim it cannot see --- IBNR needs a separate adjustment.

## Why is pricing different from reserving?

- _Pricing_ is prospective and per-policy: no claim exists yet, and we predict the frequency/severity of hypothetical future claims from policy covariates (recall the French motor lectures).
- _Reserving_ is about claims that have already happened: we condition on each claim's unfolding history — payments to date, case estimates — to predict what is still to be paid.
- Consequence for the ML setup: pricing rows are policies with static features; reserving rows are claim-quarter snapshots with _time-varying_ features. That's why this lecture needs a whole preprocessing pipeline.


# Insurance Claim Payments and Case Estimates

## The key dates

![Key dates](images/fig-event-dates.svg)

A claim is _open_ if not yet finalised.

$$
  \mathcal{O}(v) := \{ k : d_k^F > v \} \,.
$$

(Though, technically, claims sometimes finalise, and then get reopened with further payments, and refinalised again.)

## Series of incremental payments

| Event | Date | Amount |
|-------|------|--------|
| Accident | 2021-09-30 | |
| Notification | 2021-11-15 | |
| Payment #1 | 2022-05-17 | $9.00 |
| Payment #2 | 2022-08-31 | $2.00 |
| Payment #3 | 2022-10-23 | $2.00 |
| Payment #4 | 2022-12-20 | $4.00 |
| Payment #5 | 2023-02-19 | $2.00 |
| Payment #6 | 2023-05-31 | $3.00 |
| Payment #7 | 2023-07-08 | $18.00 |
| Payment #8 | 2023-09-10 | $8.00 |
| Finalisation | 2023-10-28 | |

## Case estimates

Alongside payments, the insurer holds a _case estimate_ $C_{k,v} \ge 0$: a claims assessor's live, subjective estimate of the outstanding liability at the end of quarter $v$.

The assessor is aiming for
$$
  C_{k,v} \approx O_k(v) \,.
$$

The crucial difference: $C_{k,v}$ is _known at time_ $v$, whereas $O_k(v)$ is only knowable in retrospect, once the claim finalises.

So case estimates play two roles for us:

- a _strong predictor_ to feed into the model, and
- a _benchmark to beat_ (if the model can't out-predict the assessors, why bother?).

## Payments and case estimates together

| Date | Event | Payment | Case estimate |
|------|-------|--------:|--------------:|
| 2021-11-15 | Notification, initial estimate | | $30.00 |
| 2022-05-17 | Payment #1 | $9.00 | $21.00 |
| 2022-06-20 | Review: worse than hoped | | $35.00 |
| 2022-08-31 | Payment #2 | $2.00 | $33.00 |
| 2022-10-23 | Payment #3 | $2.00 | $31.00 |
| 2022-12-20 | Payment #4 | $4.00 | $27.00 |
| 2023-02-19 | Payment #5 | $2.00 | $25.00 |
| 2023-05-31 | Payment #6 | $3.00 | $22.00 |
| 2023-07-08 | Payment #7 (settlement agreed) | $18.00 | $8.00 |
| 2023-09-10 | Payment #8 | $8.00 | $0.00 |
| 2023-10-28 | Finalisation | | $0.00 |

## From the policyholder's perspective

```{python}
#| echo: false
# One synthetic claim's life, on the nominal basis (case estimates are
# nominal). Shared prep for the two views below; uses setup.py's full-name
# imports (pandas, plt) so the case study's `import pandas as pd` slide isn't
# preempted.
claim_story_no = 7501
_claims = pandas.read_parquet("data/splice/claims.parquet")
_payments = pandas.read_parquet("data/splice/payments.parquet")
_estimates = pandas.read_parquet("data/splice/estimates.parquet")

_claim = _claims.set_index("claim_no").loc[claim_story_no]
_pay = (_payments[_payments["claim_no"] == claim_story_no]
        .sort_values("payment_time"))
_est = (_estimates[_estimates["claim_no"] == claim_story_no]
        .sort_values("txn_time"))

_ultimate = _pay["payment_inflated"].sum() / 1_000
# Plot on a "quarters since notification" axis: notification, not the accident,
# is the first moment the insurer knows of the claim, so it's the natural t=0.
# The stored times are absolute quarters, so shift everything by notification.
_t_notif_abs = _claim["occurrence_time"] + _claim["notidel"]
_pay = _pay.assign(payment_time=_pay["payment_time"] - _t_notif_abs)
_est = _est.assign(txn_time=_est["txn_time"] - _t_notif_abs)
_t_acc = -_claim["notidel"]
_t_notif = 0.0
_t_final = _claim["setldel"]

_t_paid = [_t_acc, *_pay["payment_time"], _t_final]
_cum_paid = [0.0, *(_pay["payment_inflated"].cumsum() / 1_000), _ultimate]
_t_est = [_t_notif, *_est["txn_time"], _t_final]
_case_est = [_est["OCL"].iloc[0] / 1_000, *(_est["OCL"] / 1_000), 0.0]
_incurred = [_est["incurred"].iloc[0] / 1_000,
             *(_est["incurred"] / 1_000), _ultimate]
_y_top = max(_incurred) * 1.08  # shared y-limits so the two slides overlay

# Transactions where the assessor actually issued a new estimate, as opposed
# to the automatic dollar-for-dollar knock-on from a payment (txn_type "P").
# SPLICE labels them major ("Ma"/"PMa") or minor ("Mi"/"PMi") revisions.
_major = _est["txn_type"].str.contains("Ma")
_minor = _est["txn_type"].str.contains("Mi")


def _mark_revisions(values, colour):
    for mask, marker, size, label in [(_major, "D", 14, "Major revision"),
                                      (_minor, "o", 10, "Minor revision")]:
        plt.scatter(_est.loc[mask, "txn_time"], values[mask] / 1_000,
                    marker=marker, s=size, color=colour, zorder=5,
                    label=label)


def _claim_base_plot():
    plt.figure(figsize=(5, 2.6))
    # Reserve space to the right of the axes for the legend, which sits outside
    # the plot (the claim's lines fill the frame, so an in-plot legend can't
    # avoid them).
    plt.subplots_adjust(top=0.90, left=0.13, right=0.68, bottom=0.19)
    # Labelled so "Ultimate" shows up in the legend rather than as an in-plot
    # annotation (which collided with the lines).
    plt.axhline(_ultimate, color="black", linestyle="--",
                linewidth=1, zorder=1, label="Ultimate")
    plt.ylim(0, _y_top)
    # Start at notification (t=0) and end at finalisation; the pre-notification
    # stub is bare, and setting an explicit right bound is required because
    # xlim() here (before the data is drawn) otherwise freezes autoscaling.
    plt.xlim(0, _t_final * 1.02)
    plt.xlabel("Time since notification (quarters)")
    plt.ylabel("Amount ($000s)")
```

```{python}
#| echo: false
#| fig-dpi: 350
_claim_base_plot()
plt.step(_t_paid, _cum_paid, where="post", color="#3F9999",
         linewidth=2, label="Cumulative payments")
plt.step(_t_est, _incurred, where="post", color="#CC91BC",
         linewidth=2, label="Incurred")
_mark_revisions(_est["incurred"], "#CC91BC")
plt.legend(fontsize=7, frameon=False, loc="center left",
           bbox_to_anchor=(1.0, 0.5), ncol=1, handletextpad=0.4,
           borderaxespad=0);
```

Payments step up toward the ultimate.
The _incurred_, which is cumulative payments plus the case estimate of the outstanding, is the assessor's running view of the ultimate.

## From the insurer's perspective

```{python}
#| echo: false
#| fig-dpi: 350
_claim_base_plot()
plt.step(_t_paid, [_ultimate - c for c in _cum_paid], where="post",
         color="#FF8FA9", linewidth=2, label="True outstanding")
# A hue well away from both the pink outstanding here and the mauve
# incurred on the previous slide.
plt.step(_t_est, _case_est, where="post", color="#E69F00",
         linewidth=2, label="Case estimate (OCL)")
_mark_revisions(_est["OCL"], "#E69F00")
plt.legend(fontsize=7, frameon=False, loc="center left",
           bbox_to_anchor=(1.0, 0.5), ncol=1, handletextpad=0.4,
           borderaxespad=0);
```

The true outstanding value and the _outstanding claims liability (OCL)_ (the assessor's live forecast of outstanding) both drop to zero.

# Benchmark: Aggregate Reserving

## Three levels of claim data granularity

1. _Individual transaction level_: Stores every transaction or other update for every claim. Highest granularity. Often just visible in the insured's bank statements, or in insurance company's finance records. Usually considered too granular to be useful for actuaries.
2. _Individual quarterly summaries_: This aggregates all the payments within a quarter (alternatively per month) for each claim. This is what the actuary would see.
3. _Portfolio quarterly summaries_: Aggregates all payments within a quarter (alternatively per month) across all claims. This feeds into a Chain Ladder style of reserving.

## Aggregate reserving

You have seen the classical approach to reserving already.
The insurer's past payments are aggregated into a _run-off triangle_: rows are accident years, columns are development years, and each cell holds the cumulative amount paid so far.

| Accident Year | Dev 1 | Dev 2 | Dev 3 |
|---------------|------:|------:|------:|
| 2021 | 100 | 150 | 165 |
| 2022 | 120 | 180 | ? |
| 2023 | 140 | ? | ? |

The lower-right of the triangle is the future: it hasn't happened yet, and the _reserve_ is our estimate of it.

## Chain ladder

Chain ladder fills in the triangle by assuming each accident year develops in the same proportions as the past ones.

- Development factor from Dev 1 to Dev 2:
  $f_1 = \frac{150 + 180}{100 + 120} = 1.5$
- Development factor from Dev 2 to Dev 3:
  $f_2 = \frac{165}{150} = 1.1$

Projecting forward:

| Accident Year | Dev 1 | Dev 2 | Dev 3 | Reserve |
|---------------|------:|------:|------:|--------:|
| 2021 | 100 | 150 | 165 | 0 |
| 2022 | 120 | 180 | _198_ | 18 |
| 2023 | 140 | _210_ | _231_ | 91 |

The total reserve here is $18 + 91 = 109$.

<!-- Chain ladder on the realistic portfolio lives in the case study section;
this section stays toy-sized. -->

## What chain ladder ignores

Every claim in the same accident year is treated identically.
The triangle has already thrown away:

- _who_ was injured (age, injury severity),
- _how_ the claim is progressing (legal representation, payment pattern),
- the case estimates set by claims assessors.

If two claims have paid out the same amount so far, chain ladder implicitly reserves the same for both, even if one is a minor injury about to finalise and the other is heading to court.

## Why individual claim reserving?

With claim-level data and machine learning, we can use those discarded covariates to predict each claim's outstanding amount separately.

- The portfolio reserve is then just a sum: $T(v) = \sum_{k \in \mathcal{O}(v)} O_k(v)$.
- Per-claim predictions are useful beyond the total: claims triage (which claims need senior assessors?), segmenting the reserve by injury severity or legal representation, spotting claims whose case estimates look off.
- Chain ladder remains the _benchmark_: an individual model that can't beat the triangle isn't worth its complexity.

::: {.callout-note}
Next week's guest lecture digs into _when and why_ aggregate models break down, and the trade-offs between micro- and macro-level models. Today we focus on individual-level approach and how to implement it.
:::

# Individual Claim Modelling Choices

## A few major choices upfront

1) What are the model inputs and outputs?

Predictions are made from the perspective of some fixed time $v$, the _valuation quarter_.

The individual claim model has:

- _Input_: an $X_{k,v}$ summarising the claim's history up to $v$.
- _Output_: either the ultimate claim size $U_k$ or the outstanding value $O_k(v)$.

2) Are we performing traditional regression or distributional regression?

## Two potential targets: ultimate or outstanding

The output distribution can sit on either of two quantities:

- the _ultimate_ $U_k$, or
- the _outstanding_ $O_k(v) = U_k - P_k(v)$.

```{python}
#| echo: false
# Illustrative densities for the two things a distributional model could
# forecast for one in-progress claim. The x-axis is the ultimate amount.
#   - Forecast the ultimate directly: a lognormal on [0, inf).
#   - Forecast the outstanding: a lognormal on the *outstanding*, which on the
#     ultimate axis is shifted right by the cumulative paid-to-date.
from scipy.stats import lognorm

_C_ult = "#3F9999"  # teal, as for cumulative payments earlier
_C_out = "#CC91BC"  # mauve, as for incurred earlier
_paid = 60.0        # cumulative paid to date ($000s)

# Ultimate  U ~ LogNormal;  outstanding  O ~ LogNormal (shifted by _paid).
_x = numpy.linspace(0, 260, 2000)
_pdf_ult = lognorm.pdf(_x, s=0.42, scale=95)
_xo = _x - _paid
_pdf_out = numpy.where(_xo > 0, lognorm.pdf(_xo, s=0.5, scale=48), 0.0)


def _forecast_base_plot():
    plt.figure(figsize=(5.4, 3.0))
    plt.subplots_adjust(top=0.92, left=0.145, right=0.70, bottom=0.17)
    plt.plot(_x, _pdf_ult, color=_C_ult, linewidth=2,
             label="Forecast ultimate $U_k$")
    plt.plot(_x, _pdf_out, color=_C_out, linewidth=2,
             label="Forecast outstanding $O_k(v)$")
    plt.axvline(_paid, color="black", linestyle="--", linewidth=1.2,
                label="Paid to date")
    plt.ylim(0, None)
    plt.xlim(0, 260)
    plt.xlabel("Ultimate claim amount ($000s)")
    plt.ylabel("Probability density")
    plt.legend(fontsize=7, frameon=False, loc="center left",
               bbox_to_anchor=(1.0, 0.5), handletextpad=0.5, borderaxespad=0)
```

```{python}
#| echo: false
#| fig-dpi: 350
_forecast_base_plot();
```

## Claim snapshots

![We will turn one claim into multiple "claim snapshots"/rows of input.](images/fig-snapshot-dataset.svg)

One finalised claim, viewed retrospectively from _every_ quarter it was open, yields one training row per quarter:
$$
  (X_{k,v},\ O_k(v)) \quad \text{for } v = N_k, \dots, F_k - 1 .
$$

So a claim that took $F_k - N_k$ quarters to settle contributes $F_k - N_k$ rows. The shape is what matters here: features in, one target out per row.

<!-- The full supervised dataset is
$$
  \mathcal{D} := \{(X_{k,v}, O_k(v)) : k \in \mathcal{I},\ v = N_k, \dots, F_k - 1\},
$$
where $\mathcal{I}$ is the set of _finalised_ claims --- the ones whose target $O_k(v)$ we can actually observe.
 -->


# Preprocessing One Claim

<!-- Interactive preprocessing steps, one per slide, all synced to one shared
claim (Prev/Next changes the claim everywhere). Sources in interactive-splits/;
the load/mount block lives on the first interactive slide. Clash guards:
Tailwind preflight off globally + scoped stand-in under .interactive-demo;
modules loaded as ES modules to avoid top-level const collisions. Needs
internet at view time (CDN React/Tailwind/Plotly). -->

## One claim's full history

```{=html}
<style>
/* Scoped stand-in for Tailwind preflight (which is disabled globally below,
   so it cannot restyle reveal.js/Bootstrap). */
.interactive-demo *,
.interactive-demo *::before,
.interactive-demo *::after {
  box-sizing: border-box;
  border-width: 0;
  border-style: solid;
  border-color: #e5e7eb;
}
.interactive-demo h1, .interactive-demo h2, .interactive-demo h3,
.interactive-demo h4, .interactive-demo h5, .interactive-demo h6,
.interactive-demo p, .interactive-demo figure, .interactive-demo blockquote,
.interactive-demo dl, .interactive-demo dd {
  margin: 0;
}
.interactive-demo ul, .interactive-demo ol {
  list-style: none;
  margin: 0;
  padding: 0;
}
.interactive-demo button, .interactive-demo input, .interactive-demo select,
.interactive-demo textarea {
  font-family: inherit;
  font-size: 100%;
  font-weight: inherit;
  line-height: inherit;
  color: inherit;
  margin: 0;
  padding: 0;
}
.interactive-demo button {
  background-color: transparent;
  background-image: none;
  text-transform: none;
  cursor: pointer;
}
.interactive-demo table {
  border-collapse: collapse;
  text-indent: 0;
  border-color: inherit;
  margin: 0;
}
.interactive-demo {
  font-family: -apple-system, BlinkMacSystemFont, "Segoe UI", Roboto, sans-serif;
  font-size: 16px;
  line-height: 1.4;
  text-align: left;
  max-height: 540px;
  overflow: auto;
  background: #f9fafb;
  border: 1px solid #e5e7eb;
  border-radius: 8px;
  /* Become the containing block for position:fixed descendants (the
     splitting diagram's Controls drawer) and clip them: inside reveal's
     transformed slides the "off-screen" closed drawer would otherwise sit
     visibly just right of the slide. Open, it docks inside this box. */
  contain: paint;
}
.interactive-demo svg { max-width: none; }
/* The splitting diagram centres its Train/Val/Test legend as an absolute
   overlay on the control-bar row — fine at full browser width, but at slide
   width it overlaps the buttons. Make it a normal row above them. */
#demo-splitting .pointer-events-none.absolute.inset-0 {
  position: static;
  margin-bottom: 0.5rem;
}
</style>
<div id="demo-claim" class="interactive-demo r-stretch"></div>
<script src="https://unpkg.com/react@18/umd/react.production.min.js"></script>
<script src="https://unpkg.com/react-dom@18/umd/react-dom.production.min.js"></script>
<script src="https://cdn.plot.ly/plotly-2.35.2.min.js"></script>
<script src="https://cdn.tailwindcss.com"></script>
<script>
  // Must run after the Play CDN script. Two clash guards: no global CSS
  // reset (preflight), and scope every generated utility under
  // .interactive-demo (Tailwind's "important" selector strategy). Without
  // the scoping, the Play CDN scans the WHOLE page for class names and
  // emits utilities for Bootstrap's classes too — e.g. `.collapse`
  // (visibility: collapse), which hides quarto's sidebar in the html format
  // the moment Tailwind's runtime stylesheet is injected.
  tailwind.config = {
    corePlugins: { preflight: false },
    important: ".interactive-demo",
  };
</script>
<!-- Everything below is loaded as type="module": each file gets its own
     scope (so top-level consts can't collide across files — e.g. MS_PER_DAY
     in utilities.js vs claims_splitting_diagram.js), while still publishing
     via window.*; and modules execute in document order after the page is
     parsed, so every demo <div> exists by mount time. The compiled/ files
     are generated from the .jsx sources by the pre-render esbuild step —
     edit the .jsx, never these. -->
<script type="module" src="interactive-splits/utilities.js"></script>
<script type="module" src="interactive-splits/compiled/StaticCovariates.js"></script>
<script type="module" src="interactive-splits/compiled/ContinuousTimeline.js"></script>
<script type="module" src="interactive-splits/compiled/PaymentDetails.js"></script>
<script type="module" src="interactive-splits/compiled/QuarterlyAggregation.js"></script>
<script type="module" src="interactive-splits/compiled/InflationAdjustment.js"></script>
<script type="module" src="interactive-splits/compiled/OutstandingLiabilityCalculation.js"></script>
<script type="module" src="interactive-splits/compiled/PlotlySpark.js"></script>
<script type="module" src="interactive-splits/compiled/CovariateHistorySummaries.js"></script>
<script type="module" src="interactive-splits/compiled/NeuralNetworkPreprocessing.js"></script>
<script type="module" src="interactive-splits/compiled/DevelopmentPeriodGeneration.js"></script>
<script type="module" src="interactive-splits/compiled/claims_splitting_diagram.js"></script>
<script type="module" src="interactive-splits/compiled/lecture-demos.js"></script>
```

## Noting the quarters

Each event date is binned into its calendar quarter $(A_k, N_k, F_k)$, and we also track the _development quarter_ (quarters since the accident).

```{=html}
<div id="demo-payments" class="interactive-demo r-stretch"></div>
```

::: {.callout-warning}
A side effect of discretising: a claim that opens _and_ closes within a single quarter vanishes from the snapshot dataset entirely.
:::

## Quarterly aggregation

All payments within a quarter collapse into one number, $P_{k,t}$.
Case estimates within a quarter are instead summarised by their _latest_ value.

```{=html}
<div id="demo-quarterly" class="interactive-demo r-stretch"></div>
```

## Adjusting for inflation

We need to adjust all monetary units to a common date to remove inflation.

<!-- Tie-in: SPLICE outputs each payment inflated and uninflated side by side,
so the case study can point at the two columns as this deflation step. -->

```{=html}
<div id="demo-inflation-plot" class="interactive-demo r-stretch"></div>
```

## Adjusting for inflation, applied

Each of the selected claim's quarters gets an adjustment factor from the index.

```{=html}
<div id="demo-inflation" class="interactive-demo r-stretch"></div>
```

## Calculate inflation-adjusted outstanding

Compute the outstanding $O_k(v)$ retrospectively at every quarter $v$ it was open.

```{=html}
<div id="demo-outstanding" class="interactive-demo r-stretch"></div>
```

## From history to features

Everything known about claim $k$ at valuation quarter $v$:
$$
  \mathcal{H}_{k,v} := \left\{A_k, N_k, (Z_{k,t})_{t\le v},\ (P_{k,t})_{t\le v},\ (C_{k,t})_{t\le v}\right\}
$$

Some covariates $Z_{k,t}$ are _static_ (age at accident, injury severity), others are _time-varying_ (legal representation can switch on mid-claim).

Problem: this history has a _different length_ for every $(k, v)$.
Tabular neural networks want a fixed-length vector.

The preprocessing pipeline is a deterministic map
$$
  X_{k,v} := \Phi(\mathcal{H}_{k,v}) \in \mathbb{R}^p .
$$

## Two kinds of time-varying covariates

When $\Phi$ compresses the history, the time-varying covariates split into two kinds:

1. _Only the latest value matters._ E.g. legal representation status, or the current case estimate: keep $Z_{k,v}$, discard the path. (Implicitly a _Markov-style_ assumption: the current state is a sufficient summary.)
2. _The path matters._ E.g. the payments-per-quarter series: a claim that paid $40 in one early lump differs from one that dribbled out $10 four times. These get summarised as functions of the whole history --- they are _path-dependent_ features.

<!-- TODO: align "current state" vs "path-dependent" wording with the demo,
and check for established names in the literature. (NB the survival-analysis
"external vs internal" distinction is a different thing — don't borrow it.) -->

Deciding which kind each covariate is _is itself a modelling choice_: treating the case estimate as "latest value only" throws away how often the assessor has revised it.

## Summarising the payment history

How does $\Phi$ compress a variable-length payment history into fixed columns? Summary statistics:

- development quarter $v - A_k$, notification delay $N_k - A_k$,
- count / sum / mean / std / min / max of past payments,
- the latest case estimate,
- current values of the time-varying covariates,
- the static covariates unchanged.

---

```{=html}
<div id="demo-history" class="interactive-demo r-stretch"></div>
```

## Alternative: let an RNN read the history

The claim history is naturally a _sequence_ --- one vector per development quarter:
$$
  \bigl( P_{k,t},\ C_{k,t},\ Z_{k,t} \bigr) \quad \text{for } t = N_k, \dots, v .
$$

So instead of hand-crafting summary statistics, we could:

- feed the quarter-by-quarter vectors into a recurrent layer (e.g. GRU/LSTM),
- take the final hidden state as a _learned_ summary of the history,
- concatenate the static covariates, and predict $O_k(v)$ with a dense head.

<!-- TODO: pick an RNN-reserving citation for library.bib. In our experiments
the RNN didn't beat the summary-statistics pipeline, so we keep the tabular
approach — but it's a natural design worth trying. Optional: an architecture
sketch (quarter-vectors -> RNN -> hidden state -> concat static -> dense). -->

## Preparing features for a neural network

The second half of $\Phi$ is the standard tabular recipe from earlier weeks:

- `log1p` transform dollar amounts (heavy tails!),
- one-hot encode categoricals (age band, injury severity, legal representation),
- min--max scale numeric columns.

```{=html}
<div id="demo-nn" class="interactive-demo r-stretch"></div>
```

## Creating claim snapshots

One snapshot row per development quarter, each pairing features $X_{k,v}$ with the target $O_k(v)$.

```{=html}
<div id="demo-snapshots" class="interactive-demo r-stretch"></div>
```

# Splitting Claims by Finalisation Quarter

## A portfolio of claims over time

```{python}
#| echo: false
#| fig-dpi: 350
# Companion to the "total outstanding, visually" schematic, but a sample of
# real SPLICE claims (o = notification, x = payment, X = finalisation), sorted
# by notification, no valuation line: just claims overlapping in time.
_port = _claims.copy()
_port["notif_time"] = _port["occurrence_time"] + _port["notidel"]
_port["final_time"] = _port["notif_time"] + _port["setldel"]
_port = (_port[(_port["final_time"] > 60) & (_port["final_time"] <= 100)]
         .sample(30, random_state=2026).sort_values("notif_time"))

plt.figure(figsize=(5, 3.0))
for _row, (_, _c) in enumerate(_port.iterrows()):
    _y = -_row
    plt.plot([_c["notif_time"], _c["final_time"]], [_y, _y],
             color="#3F9999", linewidth=1.2)
    _p = _payments.loc[_payments["claim_no"] == _c["claim_no"], "payment_time"]
    plt.plot(_p, [_y] * len(_p), "x", color="#3F9999", markersize=4)
    plt.plot(_c["notif_time"], _y, "o", color="#3F9999", markersize=3)
    plt.plot(_c["final_time"], _y, "X", color="#3F9999", markersize=5)
plt.yticks([])
plt.gca().spines["left"].set_visible(False)
plt.xlabel("Time (quarters)");
```

Now zoom out from one claim to the whole portfolio.
Every claim needs to be allocated to train, validation, or test --- but claims _overlap in time_, so the familiar shuffled random split deserves suspicion.

Remember, claim $k$ generates $F_k - N_k$ snapshot rows, and consecutive snapshots of the same claim are _near-duplicates_ (the history only grows by one quarter).
So: _all snapshots from one claim go to the same set_.

## The simplest approach

Order claims by finalisation quarter, then cut into _train_, _validation_, and _test_ eras:

![Those still open at $t_{\text{test}}$ are the _unfinalised_ set.](images/fig-claims-tvt-sets.svg)

## Old claims skew a naive split

Suppose we keep every claim _notified_ after the portfolio's inception --- the obvious thing to do for a _brand-new_ insurer. A hard wall then sits at the start: no claim reaches back across it.

A claim can only finalise in the _train_ era if it both _arrived_ and _closed_ early, which caps its duration at $t_{\text{train}}$. Every genuinely long-running claim is forced past the cut, into _validation_ or _test_.

```{python}
#| echo: false
#| fig-dpi: 350
# Schematic (not SPLICE data): each claim is a segment from notification (open
# circle) to finalisation (filled dot), coloured by which era its finalisation
# falls in. The hard wall on the left is the insurer's inception.
_tvt_C = {"train": "#3F9999", "val": "#E0A030", "test": "#FF8FA9",
          "old": "#B8B8B8"}
_T_TRAIN, _T_VAL, _T_ETEND = 18, 28, 36


def _tvt_era(_final):
    if _final <= _T_TRAIN:
        return "train"
    return "val" if _final <= _T_VAL else "test"


def _tvt_panel(_ax, _claims_nd, _start, _show_old):
    # _claims_nd: list of (notification, duration).
    for _row, (_n, _d) in enumerate(sorted(_claims_nd)):
        _y, _f = -_row, _n + _d
        _col = _tvt_C[_tvt_era(_f)]
        if _n < _start and _show_old:  # opened before we arrived: grey stub
            _ax.plot([_n, _start], [_y, _y], color=_tvt_C["old"], lw=1.4, zorder=2)
            _ax.plot([_start, _f], [_y, _y], color=_col, lw=1.8, zorder=3)
            _ax.plot(_n, _y, "o", mfc="white", mec=_tvt_C["old"], mew=1.0, ms=4, zorder=4)
        else:
            _ax.plot([_n, _f], [_y, _y], color=_col, lw=1.8, zorder=3)
            _ax.plot(_n, _y, "o", mfc="white", mec=_col, mew=1.1, ms=4, zorder=4)
        _ax.plot(_f, _y, "o", color=_col, ms=5, zorder=4)
    _nrows = len(_claims_nd)
    _ybot = -_nrows + 0.4
    for _x0, _x1, _c in [(_start, _T_TRAIN, "train"), (_T_TRAIN, _T_VAL, "val"),
                         (_T_VAL, _T_ETEND, "test")]:
        _ax.axvspan(_x0, _x1, color=_tvt_C[_c], alpha=0.06, zorder=0)
        _ax.annotate(_c, ((_x0 + _x1) / 2, _ybot - 0.9), ha="center", va="top",
                     fontsize=8, color=_tvt_C[_c], weight="bold")
    _ax.axvline(_start, color="black", lw=1.6, zorder=1)
    for _x, _lab in [(_T_TRAIN, r"$t_{\mathrm{train}}$"), (_T_VAL, r"$t_{\mathrm{val}}$")]:
        _ax.axvline(_x, color="black", ls="--", lw=1.0, zorder=1)
        _ax.annotate(_lab, (_x, 1.4), ha="center", va="bottom", fontsize=8)
    _ax.set_xlim(-14 if _show_old else _start - 1.5, _T_ETEND + 1.5)
    _ax.set_ylim(_ybot - 1.8, 2.6)
    _ax.set_yticks([])
    for _s in ("left", "right", "top"):
        _ax.spines[_s].set_visible(False)
    _ax.set_xlabel("Time (quarters)", fontsize=8)
    _ax.tick_params(labelsize=7)


# A handful of claims with widely varying durations; several are long runners.
_tvt_post = [(1, 3), (2, 5), (4, 24), (5, 2), (7, 27), (9, 4), (11, 20),
             (14, 3), (16, 15), (19, 7), (24, 9), (27, 3), (30, 4)]

_fig, _ax = plt.subplots(figsize=(6.2, 2.9))
_tvt_panel(_ax, _tvt_post, 0, _show_old=False)
_ax.text(0.14, 0.13, "every long claim is\nforced past $t_{\\mathrm{train}}$",
         transform=_ax.transAxes, ha="left", va="bottom", fontsize=7.5,
         color=_tvt_C["train"])
plt.tight_layout();
```

The _train_ set never sees a long claim, yet the model is judged on them --- a mismatch baked in by construction.

## Fix: arrive at a steady-state portfolio

A real reserving team never starts at inception; they inherit a portfolio already _in force_, full of claims opened years earlier. We mimic that: simulate a long _burn-in_ of history first, then keep every claim _finalised_ after a chosen start date (not every claim _notified_ after it).

```{python}
#| echo: false
#| fig-dpi: 350
# Same claims, but now we "arrive" at the portfolio start with old long-running
# claims already open (grey stubs reaching back before the start line).
_tvt_pre = [(-11, 24), (-8, 20), (-6, 12), (-4, 27)]

_fig, _ax = plt.subplots(figsize=(6.2, 3.1))
_tvt_panel(_ax, _tvt_pre + _tvt_post, 0, _show_old=True)
_ax.annotate("portfolio start", (0, 1.9), ha="center", va="bottom", fontsize=7.5)
_ax.text(0.02, 0.1, "old long-running claims,\nalready in force at the start",
         transform=_ax.transAxes, ha="left", va="bottom", fontsize=7.5,
         color="#8A8A8A")
plt.tight_layout();
```

Old long-running claims now finalise inside the _train_ era too, so every split carries the full spread of durations. We simulate many quarters and place the observation window deep in the steady state.

# Splitting Claims by Notification Quarter

## This is more complex

```{=html}
<div id="demo-splitting" class="interactive-demo r-stretch"></div>
```

## Never-finalising claims are censored

If the claim $k$ is still open at the end of our dataset, then we have not observed the ultimate.

We only know a lower bound, $U_k \ge P_k(v)$, i.e. it is a _right-censored_ observation (the survival-analysis kind).

```{python}
#| echo: false
#| fig-dpi: 350
_forecast_base_plot()
# For a censored claim we know U_k >= paid-to-date. Shade the part of the
# ultimate's density that respects this bound; the mass to the left of the
# dashed line is inadmissible and must be truncated away.
_mask = _x >= _paid
plt.fill_between(_x[_mask], _pdf_ult[_mask], color=_C_ult, alpha=0.25,
                 hatch="////", edgecolor=_C_ult, linewidth=0.0,
                 label=r"Admissible: $U_k \geq$ paid");
plt.legend(fontsize=7, frameon=False, loc="center left",
           bbox_to_anchor=(1.0, 0.5), handletextpad=0.5, borderaxespad=0);
```

The ultimate model must put _zero_ mass left of the dashed line. The outstanding model gets this free: it lives on $[0, \infty)$ from the start.

## Training directly on the censored ultimate

Keep the ultimate $U_k$ as the target and borrow the _censored likelihood_ from survival analysis.

- $\delta_k \in \{0, 1\}$: claim $k$ is _finalised_ ($\delta_k = 1$) or still _open_ ($\delta_k = 0$).
- $P_k(v) = \sum_{t \le v} P_{k,t}$: the _paid-to-date_ for claim $k$ at the training cut $v$.
- Distributional regression: $U_k \mid X_k \sim \mathcal{D}(\vartheta_k)$ with $\vartheta_k = \vartheta(X_k; \theta)$, giving density $f_k$, cdf $F_k$, and survival $S_k = 1 - F_k$.

What we actually observe:

- _Finalised_ ($\delta_k = 1$): the exact ultimate $U_k = u_k$.
- _Open_ ($\delta_k = 0$): only the bound $U_k \ge P_k(v)$ (right-censored at $P_k(v)$).

## The censored log-likelihood

A finalised claim contributes its _density_ at $u_k$; an open claim contributes its _survival_ at $P_k(v)$ (all we know is $U_k \ge P_k(v)$):

$$
  L(\theta) = \prod_{k=1}^{n} \big[\,f_k(u_k)\,\big]^{\delta_k}\,\big[\,S_k\!\big(P_k(v)\big)\,\big]^{1 - \delta_k} \,.
$$

The loss is the negative log-likelihood:

$$
  -\ell(\theta) = -\sum_{k=1}^{n} \Big[\, \delta_k \log f_k(u_k) \;+\; (1 - \delta_k)\log\big(1 - F_k(P_k(v))\big) \,\Big] \,.
$$

Open claims still inform $\theta$ through the survival term $S_k\!\big(P_k(v)\big)$: the constraint $U_k \ge P_k(v)$ is genuinely informative when the target is the ultimate.

## But if we're predicting the outstanding...

We'll predict the _outstanding_ $O_k(v)$ rather than the ultimate $U_k$ directly.

- For a censored claim, the bound $U_k \ge P_k(v)$ becomes just $O_k(v) \ge 0$.
- Any sensible model of the outstanding only predicts _positive_ values.
- So the censoring constraint is satisfied _by construction_ --- censored claims carry no usable information, and we can simply drop them.


# Case Study: Synthetic Claims

## Imports for the case study

The R side simulates the data and runs the classical benchmark:

```{r}
#| message: false
#| warning: false
library(SPLICE)
library(ChainLadder)
library(arrow)
```

The Python side does the machine learning:

```{python}
import numpy as np
import pandas as pd
from sklearn.compose import ColumnTransformer
from sklearn.linear_model import Ridge
from sklearn.metrics import root_mean_squared_error
from sklearn.pipeline import make_pipeline
from sklearn.preprocessing import FunctionTransformer, MinMaxScaler, OneHotEncoder

from keras.models import Sequential
from keras.layers import Dense
from keras.callbacks import EarlyStopping
```

## Generating synthetic claims with SPLICE

```{r}
#| eval: false
sim <- generate_data(
  n_claims_per_period = 100,   # ~10,000 claims over the full horizon
  n_periods           = 100,   # long burn-in; we observe the last ~40 quarters
  complexity          = 5,     # the most realistic scenario
  random_seed         = 20260712,
  covariates_obj      = SynthETIC::test_covariates_obj
)
claims <- cbind(sim$claim_dataset, sim$covariates_data$data)
write_parquet(claims, "data/splice/claims.parquet")
write_parquet(sim$payment_dataset, "data/splice/payments.parquet")
write_parquet(sim$incurred_dataset, "data/splice/estimates.parquet")
```

Three tables: one row per _claim_, per _payment_, and per _case estimate transaction_.

```{python}
claims = pd.read_parquet("data/splice/claims.parquet")
payments = pd.read_parquet("data/splice/payments.parquet")
estimates = pd.read_parquet("data/splice/estimates.parquet")
```

<!-- complexity 5 keeps SPLICE's default inflation, so payments come in paired
uninflated/inflated columns (payment_size / payment_inflated). -->

::: footer
SynthETIC [@avanzi2021synthetic] generates the claims and payments; SPLICE [@avanzi2023splice] adds the case estimates; the claim-level covariates plug in via SynthETIC's covariates module.
:::

## All claims {.smaller}

```{python}
#| echo: false
#| output: asis
show_df(claims)
```

## All payments {.smaller}

```{python}
#| echo: false
#| output: asis
show_df(payments)
```

## All estimates {.smaller}

```{python}
#| echo: false
#| output: asis
show_df(estimates)
```

## One synthetic claim's payments

One claim's payments (note the paired nominal/deflated columns):

```{python}
#| eval: false
payments[payments["claim_no"] == 1][...]
```

```{python}
#| echo: false
#| output: asis
show_df(payments[payments["claim_no"] == 1][["pmt_no", "payment_time", "payment_period","payment_size", "payment_inflated"]])
```

## One synthetic claim's case estimates

Its case estimate transactions (`OCL` is the case estimate):

```{python}
#| eval: false
estimates[estimates["claim_no"] == 1][...]
```

```{python}
#| echo: false
#| output: asis
show_df(estimates[estimates["claim_no"] == 1][["txn_time", "txn_type", "cumpaid", "OCL"]])
```

::: {.content-visible unless-format="revealjs"}
The package shows payments as both a nominal (`payment_inflated`) and a deflated (`payment_size`) amount, and we model the deflated ones.
The case estimates are revised at irregular times, just like the toy table earlier --- though note they are tracked in nominal dollars (assessors think in the money of the day).
:::

<!-- BUILD: claim_no 1 is short (~5 quarters, 4 payments) so both tables fit;
if a later regeneration lengthens it, use .scrollable or a shorter claim.
OCL/cumpaid are nominal (inflated) dollars. -->

## From transactions to snapshot rows

Bin each claim's key dates into quarters, and sort the claims into groups:

```{python}
# We "arrive" at the portfolio at T_START and value it at T_END: keep claims
# that finalise inside this steady-state window, not those that merely start in it.
T_START, T_END = 60, 100

claims["notif_qtr"] = np.ceil(claims["occurrence_time"] + claims["notidel"]).astype(int)
claims["final_qtr"] = np.ceil(claims["occurrence_time"] + claims["notidel"] + claims["setldel"]).astype(int)

notif, final = claims["notif_qtr"], claims["final_qtr"]
ibnr = notif > T_END
unfinalised = (notif <= T_END) & (final > T_END)
flash = (final <= T_END) & (final == notif)
observed = claims[(final > T_START) & (final <= T_END) & (final > notif)]
```

```{python}
#| echo: false
print(f"{len(claims):,} claims: {len(observed):,} finalised & usable, "
      f"{unfinalised.sum():,} unfinalised, {ibnr.sum():,} IBNR, "
      f"{flash.sum():,} settled within a quarter")
```

Every category from the theory shows up in the counts.

::: {.content-visible unless-format="revealjs"}
The simulator produced genuine IBNR claims --- occurred but not yet notified at the valuation date, so we must discard them, just as reality hides them from the insurer.
And some claims open and close within a single quarter, so they _vanish_ under the quarterly discretisation, exactly as the earlier callout warned.
:::

## Building the snapshot grid

One row per claim per open quarter, with the quarterly payments $P_{k,t}$ merged in (missing quarters become zero-payment rows).
The merge/groupby plumbing is routine (code folded away); what matters is the change of shape:

```{python}
#| code-fold: true
n_snapshots = (observed["final_qtr"] - observed["notif_qtr"]).to_numpy()
grid = observed.loc[observed.index.repeat(n_snapshots),
    ["claim_no", "occurrence_period", "notif_qtr", "final_qtr", "notidel",
     "Legal Representation", "Injury Severity", "Age of Claimant"]].copy()
grid["quarter"] = grid["notif_qtr"] + grid.groupby("claim_no").cumcount()

quarterly = (payments.rename(columns={"payment_period": "quarter"})
             .groupby(["claim_no", "quarter"])["payment_size"]
             .sum().rename("Incremental").reset_index())
grid = grid.merge(quarterly, on=["claim_no", "quarter"], how="left")
grid["Incremental"] = grid["Incremental"].fillna(0.0)
```

:::: {.columns}
::: {.column width="40%"}
Before --- one row per _claim_:

```{python}
#| echo: false
#| output: asis
show_df(observed[["claim_no", "notif_qtr", "final_qtr"]].head(3))
```
:::
::: {.column width="60%"}
After --- one row per claim per open _quarter_:

```{python}
#| echo: false
#| output: asis
show_df(grid[["claim_no", "quarter", "Incremental"]].round(2).head(6))
```
:::
::::

## History summaries, case estimates, and the target

Onto that grid we add the path-dependent features (expanding summaries of the payment history), the current-state features (latest case estimate, forward-filled between revisions), and the target (code folded away):

```{python}
#| code-fold: true
g = grid.groupby("claim_no")["Incremental"]
grid["count_Incremental"] = ((grid["Incremental"] > 0)
                             .groupby(grid["claim_no"]).cumsum())
grid["sum_Incremental"] = g.cumsum()
for stat in ["mean", "std", "min", "max"]:
    grid[f"{stat}_Incremental"] = g.expanding().agg(stat).to_numpy()
grid["std_Incremental"] = grid["std_Incremental"].fillna(0.0)

ultimate = payments.groupby("claim_no")["payment_size"].sum()
grid["Outstanding"] = grid["claim_no"].map(ultimate) - grid["sum_Incremental"]
grid["development_period"] = grid["quarter"] - grid["occurrence_period"] + 1
```

```{python}
#| code-fold: true
estimates["quarter"] = np.ceil(estimates["txn_time"]).astype(int)
latest = (estimates.sort_values("txn_time")
          .groupby(["claim_no", "quarter"])["OCL"].last()
          .rename("Estimate").reset_index())
grid = grid.merge(latest, on=["claim_no", "quarter"], how="left")
grid["Estimate"] = grid.groupby("claim_no")["Estimate"].ffill()
```

:::: {.columns}
::: {.column width="35%"}
Before --- one claim's rows:

```{python}
#| echo: false
#| output: asis
# The input columns are untouched by the transforms above (only new columns
# are added), so the "before" view can be sliced from the mutated grid.
_first_claim = grid["claim_no"].iloc[0]
show_df(grid.loc[grid["claim_no"] == _first_claim,
                 ["quarter", "Incremental"]].round(2).head(5))
```
:::
::: {.column width="65%"}
After --- summaries, estimate, and the label $O_k(v)$:

```{python}
#| echo: false
#| output: asis
show_df(grid.loc[grid["claim_no"] == _first_claim,
                 ["quarter", "sum_Incremental", "Estimate",
                  "Outstanding"]].round(2).head(5))
```
:::
::::

## The dataset after preprocessing

```{python}
#| eval: false
grid[["claim_no", "quarter", "development_period", "Incremental",
      "sum_Incremental", "Estimate", "Outstanding"]].head(8)
```

```{python}
#| echo: false
#| output: asis
show_df(grid[["claim_no", "quarter", "development_period", "Incremental",
              "sum_Incremental", "Estimate", "Outstanding"]].head(8))
```

<!-- BUILD: maybe show the full column list or a transposed .head(1) so all
covariates are visible. -->

## Splitting by finalisation quarter

```{python}
T_TRAIN, T_VAL = 88, 94  # grid already starts at T_START, so train is (60, 88]
splits = {
    "train": grid[grid["final_qtr"] <= T_TRAIN],
    "val": grid[(grid["final_qtr"] > T_TRAIN) & (grid["final_qtr"] <= T_VAL)],
    "test": grid[grid["final_qtr"] > T_VAL],
}
```

```{python}
#| echo: false
for name, df in splits.items():
    print(f"{name:6s} {len(df):7,} rows {df['claim_no'].nunique():6,} claims  "
          f"(finalisation quarters {df['final_qtr'].min()}"
          f"..{df['final_qtr'].max()})")
```

Note rows $\gg$ claims: each claim contributes one row per quarter it was open --- and every one of a claim's rows lands in the same set.

## Benchmark: chain ladder

<!-- The toy method at portfolio scale, on the _deflated_ payments (so the benchmark and the models share a basis): -->

```{r}
payments <- read_parquet("data/splice/payments.parquet")
# The triangle uses the observed window (accident quarters 61-100), shifted to
# 1-40 so it reads as ten fresh accident years valued at quarter 100.
window <- subset(payments, occurrence_period > 60 & payment_period <= 100)
window$acc_year <- ceiling((window$occurrence_period - 60) / 4)
window$dev_year <- ceiling((window$payment_period - 60) / 4) -
  window$acc_year + 1

incr <- aggregate(payment_size ~ acc_year + dev_year, window, sum)
tri <- incr2cum(as.triangle(incr, origin = "acc_year",
                            dev = "dev_year", value = "payment_size"))
round(tri / 1e6, 2)  # cumulative payments, $m
```

## The chain ladder reserve

```{r}
mack <- MackChainLadder(tri, est.sigma = "Mack")
reserve_cl <- summary(mack)$Totals["IBNR:", ]
true_outstanding <- sum(subset(payments, payment_period > 100)$payment_size)

cat(sprintf("Chain ladder reserve:   $%.1fm\n", reserve_cl / 1e6),
    sprintf("True total outstanding: $%.1fm\n", true_outstanding / 1e6))
```

As this is synthetic data, we know the future and can compare against the true value.

<!-- ::: {.content-visible unless-format="revealjs"}
How close does chain ladder get?
Compare the two printed numbers above.
The complexity-5 scenario deliberately builds in evolving claim dynamics (e.g. settlement delays that depend on the occurrence period) --- the kind of behaviour that violates chain ladder's same-development-pattern assumption, and precisely what next week's guest lecture dissects.
::: -->

<!-- BUILD: maybe add plot(mack) / cumulative development curves as an EDA
slide. NB ChainLadder labels its summary row "IBNR" but on a paid triangle
it's the outstanding-payments reserve (worth a verbal aside vs our stricter
"IBNR"). reserve_cl is read back in Python (r.reserve_cl) for the final plot. -->

## A baseline model: GLM

Before any neural network: ridge regression on the log outstanding, with the standard tabular recipe as an sklearn pipeline.

Squared-error loss on $\log O_k(v)$ is a _lognormal_ assumption for the outstanding, so every RMSE below is on the _log scale_, not in dollars.

```{python}
dollar_cols = ["Incremental", "sum_Incremental", "mean_Incremental",
               "std_Incremental", "min_Incremental", "max_Incremental"]
numeric_cols = ["development_period", "notidel", "count_Incremental"]
cat_cols = ["Legal Representation", "Injury Severity", "Age of Claimant"]

def make_glm(with_estimate):
    dollars = dollar_cols + (["Estimate"] if with_estimate else [])
    preprocess = ColumnTransformer([
        ("dollars", make_pipeline(FunctionTransformer(np.log1p),
                                  MinMaxScaler()), dollars),
        ("numeric", MinMaxScaler(), numeric_cols),
        ("categorical", OneHotEncoder(drop="first", sparse_output=False),
         cat_cols),
    ])
    return make_pipeline(preprocess, Ridge(alpha=1e-4))
```

Two versions: _without_ and _with_ the case estimate as a feature.

<!-- target is log(Outstanding): every open snapshot has Outstanding > 0, so
plain log is safe (no zero-outstanding rows in this sim). Feature dollars still
use log1p, since those can be zero. MSE on the log target = a lognormal
assumption for the outstanding, and the reported RMSEs are on the log scale.
Scalers/encoders fit inside the pipeline on train only — callback to the "fit
on train only" bullet. -->

## How do the GLMs do?

```{python}
X_train, X_test = splits["train"], splits["test"]
y_train = np.log(X_train["Outstanding"])
y_test = np.log(X_test["Outstanding"])

glms = {}
for with_est in [False, True]:
    name = f"GLM {'with' if with_est else 'without'} case estimates"
    glms[name] = make_glm(with_est).fit(X_train, y_train)
    rmse = root_mean_squared_error(y_test, glms[name].predict(X_test))
    print(f"{name}: test RMSE (log scale) = {rmse:.3f}")

rmse_ce = root_mean_squared_error(y_test, np.log(X_test["Estimate"]))
print(f"Case estimate alone: test RMSE (log scale) = {rmse_ce:.3f}")
```

Two questions for the printed numbers: does the model beat the assessors' estimates? Does giving it the case estimates help?

::: {.content-visible unless-format="revealjs"}
On this synthetic portfolio the payment history already carries most of the signal, so the case estimates may add little as a feature.
On real data, assessors know things the payment stream doesn't, and case estimates tend to matter much more.
:::

## Fitting a neural network

Swap the ridge regression for a small dense network --- same features, same target:

```{python}
#| eval: false
preprocess = make_glm(with_estimate=True)[:-1]  # the same recipe, minus the Ridge
X_train_nn = preprocess.fit_transform(X_train)
X_val_nn = preprocess.transform(splits["val"])
y_val = np.log(splits["val"]["Outstanding"])

nn = Sequential([
    Dense(30, activation="leaky_relu"),
    Dense(30, activation="leaky_relu"),
    Dense(1)])
nn.compile(optimizer="adam", loss="mse")
es = EarlyStopping(patience=10, restore_best_weights=True)
hist = nn.fit(X_train_nn, y_train, epochs=100,
              validation_data=(X_val_nn, y_val), callbacks=[es], verbose=0)
```

```{python}
#| echo: false
#| output: false
# Hidden twin of the chunk above with a model cache (the same guard idiom as
# the SPLICE generation chunk): training takes a minute, loading is instant.
# Delete models/individual-reserving-nn.keras to force a refit.
preprocess = make_glm(with_estimate=True)[:-1]
X_train_nn = preprocess.fit_transform(X_train)
X_val_nn = preprocess.transform(splits["val"])
y_val = np.log(splits["val"]["Outstanding"])

nn_path = "models/individual-reserving-nn.keras"
if Path(nn_path).exists():
    nn = keras.models.load_model(nn_path)
else:
    nn = Sequential([
        Dense(30, activation="leaky_relu"),
        Dense(30, activation="leaky_relu"),
        Dense(1)])
    nn.compile(optimizer="adam", loss="mse")
    es = EarlyStopping(patience=10, restore_best_weights=True)
    hist = nn.fit(X_train_nn, y_train, epochs=100,
                  validation_data=(X_val_nn, y_val), callbacks=[es],
                  verbose=0)
    nn.save(nn_path)
```

The validation set finally earns its keep: early stopping monitors the validation loss, and our careful split by finalisation quarter is what makes that loss trustworthy.

::: {.content-visible unless-format="revealjs"}
The paper's version of this model predicts a whole _distribution_ for each claim's outstanding (recall the distributional regression week); a point estimate of the log outstanding is enough for today's comparisons.
:::

## Evaluating the models

All three models on the test set, against the benchmark-to-beat:

```{python}
X_test_nn = preprocess.transform(X_test)
preds = {name: glm.predict(X_test) for name, glm in glms.items()}
preds["Neural network with case estimates"] = nn.predict(X_test_nn, verbose=0).flatten()
preds["Case estimates alone"] = np.log(X_test["Estimate"])

for name, pred in preds.items():
    rmse = root_mean_squared_error(y_test, pred)
    print(f"{name:34s} test RMSE (log scale) = {rmse:.3f}")
```

Per-snapshot accuracy is one lens --- but a reserving model's real job is the portfolio total.

## Predicted vs actual outstanding

```{python}
#| echo: false
#| fig-dpi: 350
plt.figure(figsize=(2.7, 2.7))
plt.scatter(y_test, preds["Neural network"], s=2, alpha=0.15,
            color="#3F9999", edgecolors="none")
add_diagonal_line()
plt.xlabel("Actual $\\log O_k(v)$")
plt.ylabel("Predicted");
```

Each point is one test-set snapshot $(X_{k,v}, O_k(v))$, plotted on the model's log scale; the dashed diagonal is perfect prediction.

## From per-claim predictions to the reserve

This is the whole point of individual reserving: predict $\widehat{O}_k(v)$ for every claim still open at the valuation date, and sum them to get $\widehat{T}(v)$.

```{python}
#| echo: false
#| output: false
# Snapshot features for claims still open at the valuation date: re-run the
# snapshot pipeline with final_qtr clamped to T_END + 1 and keep each claim's
# final row. Mirrors the "Building the snapshot grid" / "History summaries"
# chunks — keep in sync. No leakage: the join key stops at T_END.
open_claims = claims[unfinalised].copy()
open_claims["final_qtr"] = T_END + 1

n_open = (open_claims["final_qtr"] - open_claims["notif_qtr"]).to_numpy()
og = open_claims.loc[open_claims.index.repeat(n_open),
    ["claim_no", "occurrence_period", "notif_qtr", "final_qtr", "notidel",
     "Legal Representation", "Injury Severity", "Age of Claimant"]].copy()
og["quarter"] = og["notif_qtr"] + og.groupby("claim_no").cumcount()

og = og.merge(quarterly, on=["claim_no", "quarter"], how="left")
og["Incremental"] = og["Incremental"].fillna(0.0)

g_o = og.groupby("claim_no")["Incremental"]
og["count_Incremental"] = ((og["Incremental"] > 0)
                           .groupby(og["claim_no"]).cumsum())
og["sum_Incremental"] = g_o.cumsum()
for stat in ["mean", "std", "min", "max"]:
    og[f"{stat}_Incremental"] = g_o.expanding().agg(stat).to_numpy()
og["std_Incremental"] = og["std_Incremental"].fillna(0.0)
og["development_period"] = og["quarter"] - og["occurrence_period"] + 1

og = og.merge(latest, on=["claim_no", "quarter"], how="left")
og["Estimate"] = og.groupby("claim_no")["Estimate"].ffill()
X_open = og[og["quarter"] == T_END]
```

With one feature row per open claim (`X_open`, built by the same pipeline):

```{python}
reserves = {"Chain ladder": float(r.reserve_cl)}  # reticulate's R bridge
for name, glm in glms.items():
    reserves[name] = np.exp(glm.predict(X_open)).sum()
reserves["Neural network"] = float(np.exp(
    nn.predict(preprocess.transform(X_open), verbose=0)).sum())
```

## The portfolio reserves compared

```{python}
#| echo: false
#| fig-dpi: 350
future = payments[payments["payment_period"] > T_END]
true_open = future[future["claim_no"]
                   .isin(X_open["claim_no"])]["payment_size"].sum()
true_total = future["payment_size"].sum()

_short = {"GLM without case estimates": "GLM",
          "GLM with case estimates": "GLM + case est."}
_names = [_short.get(n, n) for n in reserves][::-1]
_vals = [v / 1e6 for v in list(reserves.values())[::-1]]

plt.figure(figsize=(5, 2.0))
plt.barh(_names, _vals, color="#91CCCC", height=0.55)
plt.axvline(true_open / 1e6, color="#FF8FA9", linestyle="--", linewidth=1.2)
plt.axvline(true_total / 1e6, color="black", linestyle="--", linewidth=1.2)
plt.annotate("true, visible claims", (true_open / 1e6, 1.02),
             xycoords=("data", "axes fraction"), ha="right", fontsize=7,
             color="#FF8FA9")
plt.annotate(" true, incl. IBNR", (true_total / 1e6, 1.02),
             xycoords=("data", "axes fraction"), ha="left", fontsize=7)
plt.xlim(0, max(true_total / 1e6, max(_vals)) * 1.15)
plt.xlabel("Portfolio reserve ($m)");
```

The individual models can only reserve for claims they can _see_, so their fair target is the lower value (left line); the chain ladder triangle implicitly projects the IBNR claims too, so its target is the larger value (right line).

<!-- The intended story: the individual models target the lower (visible-
claims) line, chain ladder the upper (incl. IBNR). All numbers come from the
live figure; check the narrative still holds after any regeneration. -->

# Wrapping Up

## Individual vs aggregate reserving

- _Aggregate_ (chain ladder on triangles): simple, stable, industry standard; but throws away claim-level covariates and gives no per-claim view.
- _Individual_: predicts $O_k(v)$ per open claim; the portfolio reserve is just the sum $T(v)$; supports triage and segmentation; per-claim errors partially cancel in the aggregate.
- Individual level reserving is a data engineering headache, and the total still misses IBNR, so an individual model alone is _not yet_ a full portfolio reserve.

<!-- Callback the T(v)/IBNR figure from the Introduction here. -->

## What we didn't cover

- _IBNR_: our model only reserves claims it can see; incurred-but-not-reported claims need a separate (usually aggregate) adjustment.
- _Uncertainty_: a reserve needs a risk margin, not just a point estimate; distributional forecasts and ensembling give prediction intervals.
- _When does granular beat aggregate?_

## Python package versions {.appendix data-visibility="uncounted"}

```{python}
from watermark import watermark
print(watermark(python=True, packages="keras,matplotlib,numpy,pandas,sklearn,torch"))
```

## R package versions {.appendix data-visibility="uncounted"}

```{r}
cat(sessionInfo()$R.version$version.string, "\n\n")
for (p in c("SPLICE", "SynthETIC", "ChainLadder", "arrow", "reticulate", "knitr"))
  cat(sprintf("%-12s %s\n", p, as.character(packageVersion(p))))
```

## References {.appendix data-visibility="uncounted"}

::: {#refs}
:::